Categories of Optics

### 2018-03-04 paper.tex
Some fixes

\begin{remark}
Note that the definition of the $\Optic$ category involves a coend possibly indexed by a large category. To be maximally careful, we should only ever discuss $\Optic$ categories where we know that the coend exists by some other means, e.g., by exhibiting an isomorphism with some other set. For many, but not all, of the examples given later, we provide such an isomorphism.

@@

\begin{proposition}
For any $\C$, there is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

@@

@@

\begin{proposition}
$\Optic_\C$ forms a teleological category, with the following data:
\begin{itemize}
\item The dualisable morphisms are all morphisms of the form $\iota(f, g)$;
\item The involution is given on objects by $(S, S')^* = (S', S)$, and on a morphism $\iota(f, g)$ by $\iota(g, f)$;
\item The counit $\varepsilon_{(S, S')}$  is given by the image of the symmetry $s_{S \otimes S'}$ under the isomorphism of Proposition \ref{prop-costates}.
\end{itemize}
\end{proposition}
\begin{proof}
That morphisms of the form $\iota(f, g)$ constitute a category and that $\iota$ is a symmetric monoidal involution is clear.

@@

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a lens, i.e. that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ for some $\phi : M \to M$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$ on the nose.

@@

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has as a representative $\langle [\id_s, \fget], \fput \rangle$.

@@

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

>>>>

\begin{remark}
Note that the definition of the $\Optic$ category involves a coend possibly indexed by a large category. To be maximally careful, we should only ever discuss $\Optic$ categories where we know that the coend exists by some other means, e.g., by exhibiting an isomorphism with some other set. For many---but not all---of the examples given later, we provide such an isomorphism.

@@

\begin{proposition}
There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

@@

\begin{remark}
This functor is not necessarily faithful, see Remark \ref{lens-iota-not-faithful}
\end{remark}

@@

\begin{proposition}
$\Optic_\C$ forms a teleological category, with the following data:
\begin{itemize}
\item The dualisable morphisms are all morphisms of the form $\iota(f, g)$;
\item The involution is given on objects by $(S, S')^* = (S', S)$, and on a morphism $\iota(f, g)$ by $\iota(g, f)$;
\item The counit $\varepsilon_{(S, S')}$  is given by the image of the symmetry $s_{S \otimes S'}$ under the isomorphism of Proposition \ref{prop-costates}. \todo{It is a little annoying that a twist is necessary, but it has to go somewhere. The other option is in the definition of $\otimes$ but that seems worse.}
\end{itemize}
\end{proposition}
\begin{proof}
That morphisms of the form $\iota(f, g)$ constitute a category and that $\iota$ is a symmetric monoidal involution is clear. \todo{actually check monoidal part}

@@

\begin{remark} \label{lens-iota-not-faithful}
For the category of sets, the functor $\iota(-, -) : \Set \times \Set^\op \to \Optic_\times$ is not faithful. The problem is the empty set: the functor $0 \times (-)$ is not faithful. Any pair of maps $f : 0 \to A$, $g : A' \to S'$ yield equivalent optics $\iota(f, g)$, as the corresponding $\fget$ and $\fput$ functions must be the unique maps from $0$.

\todo{Show that for the category of nonempty sets things .}
\end{remark}

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a lawful lens in our sense, i.e. that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ for some $\phi : M \to M$. Then $\fput$ and $\fget$ for this lens obey the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$ on the nose.

@@

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lawful lens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has as a representative $\langle [\id_s, \fget], \fput \rangle$.

@@

\begin{proposition}
Suppose $\C$ has pullbacks and that there is a morphism $x : 1 \to A$. If $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}
